Comparison of Spatial Partitioning Strategies

### 2025-02-27 paper/paper.tex
paper: Wrote some analysis parts.

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{svg}

@@

\section{Introduction}
When I was developing a small physics engine, I encountered performance issues
during the  collision detection phase. In this phase the program figures out
which objects are close enough to be touching and overlapping. This wouldn't be
an issue in a small simulation with few objects, but I was striving for
something grand running in real time which made every millisecond count. Due to
these constraints I had to find a way to speed things up.

On my optimization journey I encountered the concept of spatial partitioning -
an elegant solution to my problem. Diving deeper I discovered the three
techniques I'll be comparing in this paper, along side the naive solution. The
basic theory of the technique is simply to group objects close to each other to
avoid checking two objects on the opposite side of the world. This drastically
decreases the number of collision checks performed each frame.

As a consequence of this constraint I had to find a way to optimize this phase.
That is what sent me down the rabbit hole of spatial partitioning as an
optimization to do query objects close to each other in space. During my
research I found out about the three strategies I'm comparing in this paper. I'm
doing this paper within the programming course.

\section{Theory}
\subsection{Spatial Partitioning}
Spatial partitioning is an optimization technique used to make programs run
faster. It's often found within physics applications. The theory behind this
technique is to reduce the number of times one object is checked against
another. Suppose we have one hundred objects in our world. The naive way to
check each one of these against the others results in a ten thousand checks.
Since checks are expensive to make, we want to perform as few of them as
possible.

\subsection{Quadtree}
A quadtree is a recursive data structure, meaning it contains pointers to each
objects of the same type as itself. It's in the family of tree data structures.
The data structure is comprised of nodes. Each node has four pointers to other
nodes, could also be referred to as branches or leaves. This strategy works by
inserting points into a leaf until the leaf reaches its capacity, then it splits
into four equally large sub leaves. This subdivision pattern is shown in figure
\ref{fig:quadtree}.

@@

\subsection{The Stack and Heap}
A program uses two different types of memory, the stack and the heap. When the
program starts a fixed amount of bytes are allocated for the program to use,
this is what is called the stack. When the program creates a variable it is
allocated on the stack, which is done by incrementing the stack pointer by the
size of the variable data type. Since a stack allocation is only a single
instruction it is incredibly fast. The heap is much larger than the stack but
slower because you have to request space within it from the operating system.
Allocating is considered to be a slow operation since there's a lot of overhead
introduced.

@@

\subsection{Static Grid}
The static grid is built using a simple array. Indexing into the array is an
$O(1)$ operation. The caveat with this approach is you're forced to know how
big the area you want the to partition to be before compiling the program. Both
insertion and querying only requires a simple index into the grid array which
makes this approach super fast.

\subsection{Spatial Hashing Grid}
The spatial hashing grid is built using the same technique as a hash map. The
method I used to write my spatial hashing grid was to allocate an arbitrarily
sized array. When inserting into the structure I first hash the coordinate of
the object, then run a modulo operation on it giving me a number within the
array bounds, and then it is inserted into the cell.

@@

\section{Results}
This section presents the results of the benchmarking done on the three
strategies - quadtrees, spatial hashing and static grids. The metrics measured
was average runtime: the sum of all operations needed to run the program.
Insertions: how long it took to insert all the objects into the data structure.
Collision testing: total time spent checking objects against each other, which
includes querying. Querying: average time spent retrieving objects from the data
structure. Clearing: how long it takes to reset the data structure making it
ready for rebuilding. Each metric was measured with a range of object count,
labeled as 'box count' in the graphs, from ten to 2500 with an increase of ten
objects between each measurement. At each stage was run 32 times and the results
were averaged to give smoother graphs. Two benchmarks was run, one with an even
distribution of objects and one with an uneven distribution.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{even-naive.png}
    \caption{Runtime of naive solution and partitioning strategies.}
    \label{fig:naive}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{even-average.png}
    \caption{Average runtime of spatial partitioning strategies with evenly
    distributed objects.}
    \label{fig:even-average}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
Figure \ref{fig:uneven-average} presents the average runtime for the program
using the different strategies with an uneven object distribution. This includes
all other measuring metrics. As the object count increases the spatial hashing
trends towards a $O(n^2)$ runtime complexity.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{uneven-average.png}
    \caption{Average runtime of spatial partitioning strategies with unevenly
    distributed objects.}
    \label{fig:uneven-average}
\end{figure}

@@

\subsubsection{Evenly Distributed Objects}
The runtime complexity of the insertion algorithms for evenly distributed
objects is shown in figure \ref{fig:even-insert}.
\begin{figure}[h]
    \centering
    \caption{Average insertion time per object with evenly distributed objects.}
    \includegraphics[width=0.5\textwidth]{even-insert.png}
    \label{fig:even-insert}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
The runtime complexity of the insertion algorithms for evenly distributed
objects is shown in figure \ref{fig:uneven-insert}.
\begin{figure}[h]
    \centering
    \caption{Average insertion time per object with unevenly distributed
    objects.}
    \includegraphics[width=0.5\textwidth]{uneven-insert.png}
    \label{fig:uneven-insert}
\end{figure}

@@

\subsubsection{Evenly Distributed Objects}
Figure \ref{fig:even-collision} shows the total time spent checking for
colliding objects per iteration with an even distribution. This time also
includes the time taken for querying.
\begin{figure}[h]
    \centering
    \caption{Average collision time for all objects with evenly distributed
    objects.}
    \includegraphics[width=0.5\textwidth]{even-collision.png}
    \label{fig:even-collision}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
Figure \ref{fig:uneven-collision} shows the total time spent checking for
colliding objects per iteration with an uneven distribution. This time also
includes the time taken for querying.
\begin{figure}[h]
    \centering
    \caption{Average collision time for all objects with unevenly distributed
    objects.}
    \includegraphics[width=0.5\textwidth]{uneven-collision.png}
    \label{fig:uneven-collision}
\end{figure}

\subsubsection{Evenly Distributed Objects}
Figure \ref{fig:even-query} represents the average time taken to query the space
per object with an even distribution.
\begin{figure}[h]
    \centering
    \caption{Average query time per object with evenly distributed objects.}
    \includegraphics[width=0.5\textwidth]{even-query.png}
    \label{fig:even-query}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
Figure \ref{fig:uneven-query} represents the average time taken to query the space
per object with an uneven distribution.
\begin{figure}[h]
    \centering
    \caption{Average query time per object with unevenly distributed objects.}
    \includegraphics[width=0.5\textwidth]{uneven-query.png}
    \label{fig:uneven-query}
\end{figure}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{svg}
\usepackage{float}

@@

\section{Introduction}
When I was developing a small physics engine, I encountered performance issues
during the  collision detection phase. In this phase the program figures out
which objects are close enough to be touching and overlapping. This would not be
an issue in a small simulation with few objects, but I was striving for
something grand running in real time which made every millisecond count. Due to
these constraints I had to find a way to speed things up.

On my optimization journey I encountered the concept of spatial partitioning -
an elegant solution to my problem. Diving deeper I discovered the three
techniques I will be comparing in this paper, along side the naive solution. The
basic theory of the technique is simply to group objects close to each other to
avoid checking two objects on the opposite side of the world. This drastically
decreases the number of collision checks performed each frame.

As a consequence of this constraint I had to find a way to optimize this phase.
That is what sent me down the rabbit hole of spatial partitioning as an
optimization to do query objects close to each other in space. During my
research I found out about the three strategies I am comparing in this paper. I
am doing this paper within the programming course.

\section{Purpose \& Scope}
The purpose of this paper is to find the most optimal strategy for dividing up
space to improve the speed of spatial querying of nearby objects.

This study focuses mainly on CPU performance and runtime in exchange for less
memory efficient implementation. Therefore time based evaluations of insertion,
query time, clearing time and general collision testing time will be performed.

The strategies are going to be evaluated on time based metrics - insertion,
query, clearing and time spend doing collision testing. These evaluations are
going to take place with two different distribution strategies of the objects
though the same seed will be used for all tests so the pseudo-random numbers
stay the same.

These strategies could be extended into the third dimension but this study
focuses solely focused on two dimensional space.

The questions I'm going to answer are:
\begin{itemize}
    \item At what point is it worth implementing a more advanced data structure?
    \item In what use case are the different algorithms most useful?
\end{itemize}

\section{Theory}
\subsection{Spatial Partitioning}
Spatial partitioning is an optimization technique used to make programs run
faster. It is often found within physics applications. The theory behind this
technique is to reduce the number of times one object is checked against
another. Suppose we have one hundred objects in our world. The naive way to
check each one of these against the others results in a ten thousand checks.
Since checks are expensive to make, we want to perform as few of them as
possible.

\subsection{Quadtree}
A quadtree is a recursive data structure, meaning it contains pointers to each
objects of the same type as itself. It is in the family of tree data structures.
The data structure is comprised of nodes. Each node has four pointers to other
nodes, could also be referred to as branches or leaves. This strategy works by
inserting points into a leaf until the leaf reaches its capacity, then it splits
into four equally large sub leaves. This subdivision pattern is shown in figure
\ref{fig:quadtree}.

@@

\subsection{The Stack and Heap}
A program uses two different types of memory, the stack and the heap. When the
program starts a fixed amount of bytes are allocated for the program to use,
this is what is called the stack. When the program creates a variable it is
allocated on the stack, which is done by incrementing the stack pointer by the
size of the variable data type. Since a stack allocation is only a single
instruction it is incredibly fast. The heap is much larger than the stack but
slower because you have to request space within it from the operating system.
Allocating is considered to be a slow operation since there is a lot of overhead
introduced.

@@

\subsection{Static Grid}
The static grid is built using a simple array. Indexing into the array is an
$O(1)$ operation. The caveat with this approach is you are forced to know how
big the area you want the to partition to be before compiling the program. Both
insertion and querying only requires a simple index into the grid array which
makes this approach super fast.

\subsection{Hash Map}
A hash map is a data structure which key value pairs allowing for aliasing of
values. For example, a hash map using strings as keys and integers as values
could map the key "life" to the value 42, meaning a query into the hash map for
"life" would return 42. The way this works is by using a cryptographic hashing
function to generate a unique hash for the object. This hash is a simple number
which is usually very large. To make the hash useful you run modulo on it to
make sure it's between zero and the number of buckets inside the hash map. The
same method is used for retrieval, first running the key through the same
hashing function, then wrapping it using modulo and returning what lies in the
bucket. The only caveat with this is there could be collisions within the hash
map. Two hashes could wrap into the same index. There are a multitude of ways to
handle collisions like open addressing and separate chaining.

\subsection{Spatial Hashing Grid}
The spatial hashing grid is built using the same technique as a hash map. The
method I used to write my spatial hashing grid was to allocate an arbitrarily
sized array. When inserting into the structure I first hash the coordinate of
the object, then run a modulo operation on it giving me a number within the
array bounds, and then it is inserted into the cell. No collision resolution is
taking place inside the structure meaning two cells could have the same bucket
index which means queries for the one of the cells returns the objects in the
other as well.

@@

\section{Results \& Analysis}
This section presents the results and offers an analysis of the benchmarking
done on the three strategies - quadtrees, spatial hashing and static grids. The
metrics measured was average runtime: the sum of all operations needed to run
the program. Insertions: how long it took to insert all the objects into the
data structure. Collision testing: total time spent checking objects against
each other, which includes querying. Querying: average time spent retrieving
objects from the data structure. Clearing: how long it takes to reset the data
structure making it ready for rebuilding. Each metric was measured with a range
of object count, labeled as 'box count' in the graphs, from ten to 2500 with an
increase of ten objects between each measurement. At each stage was run 32 times
and the results were averaged to give smoother graphs. Two benchmarks was run,
one with an even distribution of objects and one with an uneven distribution.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{even-naive.png}
    \caption{Runtime of naive solution and partitioning strategies.}
    \label{fig:naive}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{even-average.png}
    \caption{Average runtime of spatial partitioning strategies with evenly
    distributed objects.}
    \label{fig:even-average}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
Figure \ref{fig:uneven-average} presents the average runtime for the program
using the different strategies with an uneven object distribution. This includes
all other measuring metrics. As the object count increases the spatial hashing
trends towards a $O(n^2)$ runtime complexity.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{uneven-average.png}
    \caption{Average runtime of spatial partitioning strategies with unevenly
    distributed objects.}
    \label{fig:uneven-average}
\end{figure}

\subsubsection{Analysis}

With evenly distributed objects The three strategies perform very similarly,
though, spatial hashing does start showing more of an exponential runtime as the
object count increases. Could be due to the implementation of the spatial
hashing. Since it is using a hash map as its underlying data structure two
different cells could have the same index within the same hash map meaning
collision checks are happening for objects which are not close to each other in
space. This happens because there is a limited amount of buckets available to
store objects inside. The result of this scarcity is two hashes wrapping to the
same bucket index. As we can see in \ref{fig:uneven-average} this issue becomes
more apparent as the objects are less spread out.

@@

\subsubsection{Evenly Distributed Objects}
The runtime complexity of the insertion algorithms for evenly distributed
objects is shown in figure \ref{fig:even-insert}. The metric being times is the
insertion of all objects.
\begin{figure}[H]
    \centering
    \caption{Average insertion time per object with evenly distributed objects.}
    \includegraphics[width=0.5\textwidth]{even-insert.png}
    \label{fig:even-insert}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
The runtime complexity of the insertion algorithms for evenly distributed
objects is shown in figure \ref{fig:uneven-insert}. The metric being times is
the insertion of all objects.
\begin{figure}[H]
    \centering
    \caption{Average insertion time per object with unevenly distributed
    objects.}
    \includegraphics[width=0.5\textwidth]{uneven-insert.png}
    \label{fig:uneven-insert}
\end{figure}

\subsubsection{Analysis}
There's virtually no difference in performance between the even and uneven
insertion times. This is because the codepath taken for each insertion is the
same, save for a different hash and bucket index. Because there is only a fixed
amount of boxes able to be stored in each grid cell in both grid types the
insertion time is run in $O(1)$, constant time. The case isn't the same for the
quadtree since its a recursive algorithm. Each node has four sub-nodes which get
traversed each insertion leading to a $O(\log_4 n)$ runtime since each step in
the algorithm eliminates three of the sub-nodes. Once the correct node has been
found the insertion time is a constant $O(1)$ since it uses the same technique
as the other two.

\subsubsection{Evenly Distributed Objects}
Figure \ref{fig:even-query} represents the average time taken to query the space
per object with an even distribution.
\begin{figure}[H]
    \centering
    \caption{Average query time per object with evenly distributed objects.}
    \includegraphics[width=0.5\textwidth]{even-query.png}
    \label{fig:even-query}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
Figure \ref{fig:uneven-query} represents the average time taken to query the space
per object with an uneven distribution.
\begin{figure}[H]
    \centering
    \caption{Average query time per object with unevenly distributed objects.}
    \includegraphics[width=0.5\textwidth]{uneven-query.png}
    \label{fig:uneven-query}
\end{figure}

\subsubsection{Analysis}

@@

\subsubsection{Evenly Distributed Objects}
Figure \ref{fig:even-collision} shows the total time spent checking for
colliding objects per iteration with an even distribution. This time also
includes the time taken for querying.
\begin{figure}[H]
    \centering
    \caption{Average collision time for all objects with evenly distributed
    objects.}
    \includegraphics[width=0.5\textwidth]{even-collision.png}
    \label{fig:even-collision}
\end{figure}

\subsubsection{Unevenly Distributed Objects}
Figure \ref{fig:uneven-collision} shows the total time spent checking for
colliding objects per iteration with an uneven distribution. This time also
includes the time taken for querying.
\begin{figure}[H]
    \centering
    \caption{Average collision time for all objects with unevenly distributed
    objects.}
    \includegraphics[width=0.5\textwidth]{uneven-collision.png}
    \label{fig:uneven-collision}
\end{figure}

\section{Discussion}

\section{Sources}
Nystrom, Robert (2011) \textit{Game Programming Patterns}
